The promise of community-driven preprints in ecology and evolution

### 2024-10-04 ms/ms.qmd
Dealing with more comments

# Abstract
Publishing preprints is quickly becoming commonplace in ecology and evolutionary biology. Preprints can facilitate the rapid sharing of scientific knowledge establishing precedence and enabling feedback from the research community before peer review. Yet, significant barriers to preprint use exist including language barriers, a lack of understanding about the benefits of preprints and a lack of diversity in the types of research outputs accepted (e.g., reports). Community driven preprint initiatives can allow a research community to come together to break down these barriers to improve equity and coverage of global knowledge. Here, we explore the first preprints uploaded to EcoEvoRxiv, a community-driven preprint server for ecologists and evolutionary biologists, to characterise preprint practices in ecology, evolution and conservation. Our perspective piece highlights some of the unique initiatives that EcoEvoRxiv has taken to break down barriers to scientific publishing by exploring the composition of articles, how gender and career stage influence preprint use, whether preprints are associated with greater open science principles (e.g., code and data sharing), and tracking preprint publication outcomes. Our analysis identifies areas that we still need to improve upon but highlight how community-driven initiatives, such as EcoEvoRxiv, can play a crucial role in shaping publishing practices in biology.

@@

*EcoEvoRxiv* is run by the Society for Open, Reliable, and Transparent Ecology and Evolutionary Biology (SORTEE)[@o2021towards]. Originally launched in 2018 on the Center for Open Science preprint platform, *EcoEvoRxiv* has become a popular preprint server for ecologists and evolutionary biologists. The server has since been adopted by the California Digital Library (CDL). Editors are ecologists and evolutionary biologists from across the globe who volunteer their time to screen papers and push new initiatives in the preprint space. *EcoEvoRxiv* allows authors to post both preprints and postprints (also known as author-accepted manuscripts). While preprints are versions of manuscripts posted by authors before peer-review, postprints are versions of peer-reviewed and accepted articles but without typesetting and formatting by a journal. The main reason for publishing postprints on a preprint server is to ensure published articles are openly accessible to everyone without a paywall (i.e., green open access). Postprints can be published anytime, provided that journals allow it (which most do; see https://www.sherpa.ac.uk/romeo/).

```{r, fig-summary}
#| label: fig-summary
#| fig-cap: "Summary of submissions to *EcoEvoRxiv*. A) Number of submissions (preprints and postprints) published on *EcoEvoRxiv* between 2018 and 2023. *EcoEvoRxiv* was established in June 2018 before the launch in November 2018. Notable milestones include *EcoEvoRxiv* transitioning to the California Digital Library (CDL), the acceptance of submissions in Spanish and Portuguese, and the acceptance of the first IUCN Red List Ecosystem report; B) Geographic origin of the preprints and postprints uploaded to *EcoEvoRxiv*, inferred from the country of affiliation of the submitting author; C) Taxa used/covered in the articles posted to *EcoEvoRxiv* (n = 1080 articles); D) Types of preprints accepted on *EcoEvoRxiv* (n = 620 articles). E) Academic age of authors submitting to *EcoEvoRxiv* along with the gender of the submitting author. Values lower than zero are indicative of authors who uploaded preprints before their first scientific publication in a journal. Map base source:  R Package "maps" v.3.4.2. Shapefile: Natural Earth https://www.naturalearthdata.com/about/terms-of-use/."

@@

To better understand preprint (and postprint) use on *EcoEvoRxiv*, we downloaded metadata on the accepted submissions in EcoEvoRxiv as of  `r gsub("UTC", "", dates[2])` (see Supplement for more details on methods). We consider both preprints and postprints as ‘submissions’. After removing five duplicate titles – suggesting that a few authors created multiple submissions for the same preprint rather than updating the existing submission – we had data for a total of `r pr` articles with ~55--60 submissions published approximately monthly in the last two years ([@fig-summary]A).  *EcoEvoRxiv* hosts submissions from authors based in `r length(unique(countries$country))` countries, with 90% coming from just `r nrow(most_use)-1` countries. North America, Australia, and European countries upload the most preprints, with many fewer coming from countries in Africa, Central America, and parts of Asia ([@fig-summary]B).  Submissions covered all major taxonomic groups, with the most common groups being vertebrates (`r sum_taxa$perc[6]`%), plants (`r sum_taxa$perc[1]`%), and invertebrates (`r sum_taxa$perc[5]`%) ([@fig-summary]C).

@@

Accepting a greater diversity of article types allows *EcoEvoRxiv* to help deal with the 'grey literature' problem, whereby data that are relevant for research syntheses are not published in typical peer-reviewed journals [@haddaway2015shades; @haddaway2020eight]. *EcoEvoRxiv* has made a concerted effort to diversify the types of submissions accepted. This is reflected by `r nt`% of the submissions on *EcoEvoRxiv* being books, book chapters, reports, and other research output types, which are typically considered ‘grey literature’ in ecology and evolutionary biology. As a result, submissions to *EcoEvoRxiv* are more diverse than those on preprint servers which have more restrictive submission policies. For example, *bioRxiv* only accepts research articles (https://www.biorxiv.org/submit-a-manuscript).

Typical research articles are still the most common type of submission to *EcoEvoRxiv* (`r sum_types[sum_types$type == "research article",]$perc`%), followed by reviews and meta-analyses (`r sum_types[sum_types$type == "reviews and meta-analyses",]$perc`%) and opinion papers (`r sum_types[sum_types$type == "opinion",]$perc`%) ([@fig-summary]D). Currently, *EcoEvoRxiv* does not host many reports, particularly from government or industry, but has formed fruitful partnerships with the International Union for Conservation of Nature (IUCN). For example, IUCN Red-list Ecosystem Reports are now posted to *EcoEvoRxiv* and our community has been able to work closely with the IUCN to ensure these documents meet the IUCN requirements. We would encourage authors to consider posting books, book chapters, and reports to ensure that they are openly accessible and more easily found. Accepted *EcoEvoRxiv* submissions are given a unique DOI and are indexed on Google Scholar.

@@

Research can take a while to be published (see below). Early and Mid-Career Researchers (EMCRs) (~10 years post-PhD) are under pressure to publish rapidly to be competitive in job applications, promotions, and obtaining grants to progress their careers [@vale2015accelerating; @sarabipour2019value]. Preprints are especially useful for EMCRs because they can achieve faster dissemination and greater visibility [@fu2019releasing]. As such, EMCRs may be expected to make use of preprints more than colleagues at later career stages. We collected data on the 'academic age' of submitting authors by looking at Google Scholar profiles of authors (when available) and recording their first year of publication in a peer-reviewed journal. While this is a rough estimate of career stage, there was evidence that the number of preprints posted decreases with later career stages (negative binomial glm: year slope = `r res[2, "Estimate"]`, SE: `r round(res[2, "Std. Error"], 2)`, *p* < 0.001, n = `r nrow(count_dat[complete.cases(count_dat),])` years). Most preprints were submitted by authors who published their first paper in the last ~10 years ([@fig-summary]E), with the median year since first publication being `r median(data2$submitting_author_first_publication_year, na.rm = TRUE)` (mean = `r mean(data2$submitting_author_first_publication_year, na.rm = TRUE)`; SD = `r sd(data2$submitting_author_first_publication_year, na.rm = TRUE)`, n = `r sum(complete.cases(data2$submitting_author_first_publication_year))`). These patterns support the expectation that EMCRs may use preprints to make their work more visible and disseminate their findings more quickly. However, we acknowledge that validating this conclusion does require more rigorous experimental approaches.

Gender differences in preprint use and publication outcomes have also been observed in several research fields, including ecology and evolutionary biology [@fox2019gender; @wehner2020comparison]. Therefore, such discrepancies are expected to manifest in preprint use on *EcoEvoRxiv*, but it is unclear to what extent. Understanding gender publishing patterns is challenging with observational data such as ours because we cannot know the gender of authors for certain, but we can use a data-driven approach to ascertain the probability that a particular name is of a given gender (man or woman). To obtain a rough idea of an author's gender, we used the R package *gender* (v.`r utils::packageVersion("gender")`; @mullen2021predict) to predict the most likely gender of the submitting author of a preprint. We only used algorithm-assigned names where the gender was identified with 95% certainty. For the remaining names, we performed manual searches to determine gender based on the pronouns and photographs from professional and personal websites. We acknowledge that our approach does not capture self-assigned and non-binary genders. As such, our assumptions about an author's gender identity may be incorrect. Our data on gender had only two missing values--one where the first name of the submitting author was missing and the other one for a collective submission. As expected, we found that women were less likely to submit to *EcoEvoRxiv* compared to men (women: `r (table(data4$gender)[1] / sum(table(data4$gender))*100)`%), which may reflect the broader publishing disparities between male and female scientists [@fox2019gender].

# 3. Following the journey of a preprint on *EcoEvoRxiv*: from submission to publication
```{r, fig-pubsummary}
#| label: fig-pubsummary
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Summary of the publication status of articles on *EcoEvoRxiv*. A) Time between uploading an preprint to *EcoEvoRxiv* and its publication as a peer-reviewed journal article. Articles that were published within 2 months (60 days) of being submitted are considered postprints and are not included in this figure. B) Access status of articles linked to *EcoEvoRxiv* preprints classified as “Open Access” or “Not Open Access”. “Unknown” status is for articles not explicitly identified as being open access or not on the Unpaywall platform. C) Sub-types of open access status of articles linked to *EcoEvoRxiv* preprints. For full details on the meaning of each category see  https://support.unpaywall.org/support/solutions/articles/44001777288-what-do-the-types-of-oa-status-green-gold-hybrid-and-bronze-mean-.

@@

We estimated how long it takes to publish a research paper in ecology and evolution by recording the time between initial submission to *EcoEvoRxiv*, and its final acceptance in a peer-reviewed journal. In total, `r unpub` submissions remained unpublished (`r (unpub / nrow(data2))*100`%, n = `r nrow(data2)`) at the time when these data were collected. Not all of these submissions, however, are anticipated to be published in a peer-reviewed journal (e.g., reports). Nonetheless, the median time to publication for preprints was `r sum_preprints$median` days (8 months) (mean = `r sum_preprints$mean`; SD = `r sum_preprints$sd` days) with the maximum time to publication being `r sum_preprints$max` days or `r sum_preprints$max/365` years ([@fig-pubsummary]A). Our results largely confirm the extended timeframes that most authors experience between writing their research papers and their publication.

>>>>

# Abstract
Publishing preprints is quickly becoming commonplace in ecology and evolutionary biology. Preprints can facilitate the rapid sharing of scientific knowledge establishing precedence and enabling feedback from the research community before peer review. Yet, significant barriers to preprint use exist including language barriers, a lack of understanding about the benefits of preprints and a lack of diversity in the types of research outputs accepted (e.g., reports). Community driven preprint initiatives can allow a research community to come together to break down these barriers to improve equity and coverage of global knowledge. Here, we explore the first preprints uploaded to *EcoEvoRxiv*, a community-driven preprint server for ecologists and evolutionary biologists, to characterise preprint practices in ecology, evolution and conservation. Our perspective piece highlights some of the unique initiatives that *EcoEvoRxiv* has taken to break down barriers to scientific publishing by exploring the composition of articles, how gender and career stage influence preprint use, whether preprints are associated with greater open science principles (e.g., code and data sharing), and tracking preprint publication outcomes. Our analysis identifies areas that we still need to improve upon but highlight how community-driven initiatives, such as *EcoEvoRxiv*, can play a crucial role in shaping publishing practices in biology.

@@

*EcoEvoRxiv* is run by the Society for Open, Reliable, and Transparent Ecology and Evolutionary Biology (SORTEE)[@o2021towards]. Originally launched in 2018 on the Center for Open Science preprint platform, *EcoEvoRxiv* has become a popular preprint server for ecologists and evolutionary biologists. The server has since been adopted by the California Digital Library (CDL). Editors are ecologists and evolutionary biologists from across the globe who volunteer their time to screen papers and push new initiatives in the preprint space. *EcoEvoRxiv* allows authors to submit both preprints and postprints (also known as author-accepted manuscripts). While preprints are versions of manuscripts posted by authors before peer-review, postprints are versions of peer-reviewed and accepted articles but without typesetting and formatting by a journal. The main reason for publishing postprints on a preprint server is to ensure published articles are openly accessible to everyone without a paywall (i.e., green open access). Postprints can be published anytime, provided that journals allow it (which most do; see https://www.sherpa.ac.uk/romeo/).

```{r, fig-summary}
#| label: fig-summary
#| fig-cap: "Summary of articles posted to *EcoEvoRxiv*. A) Number of articles (preprints and postprints) published on *EcoEvoRxiv* between 2018 and 2023. *EcoEvoRxiv* was established in June 2018 before the launch in November 2018. Notable milestones include *EcoEvoRxiv* transitioning to the California Digital Library (CDL), the acceptance of preprints and postprints in Spanish and Portuguese, and the acceptance of the first IUCN Red List Ecosystem report; B) Geographic origin of the preprints and postprints uploaded to *EcoEvoRxiv*, inferred from the country of affiliation of the submitting author; C) Taxa used/covered in the articles posted to *EcoEvoRxiv* (n = 1080 articles); D) Types of preprints accepted on *EcoEvoRxiv* (n = 620 articles). E) Academic age of authors posting to *EcoEvoRxiv* along with the gender of the submitting author. Values lower than zero are indicative of authors who uploaded preprints before their first scientific publication in a journal. Map base source:  R Package "maps" v.3.4.2. Shapefile: Natural Earth https://www.naturalearthdata.com/about/terms-of-use/."

@@

To better understand preprint (and postprint) use on *EcoEvoRxiv*, we downloaded metadata on the accepted articles available on *EcoEvoRxiv* as of  `r gsub("UTC", "", dates[2])` (see Supplement for more details on methods). We consider both preprints and postprints as ‘articles’. After removing five duplicate titles – suggesting that a few authors uploaded the article as separate submissions rather than updating the existing article – we had data for a total of `r pr` articles with ~55--60 posted approximately monthly in the last two years ([@fig-summary]A).  *EcoEvoRxiv* hosts articles from authors based in `r length(unique(countries$country))` countries, with 90% coming from just `r nrow(most_use)-1` countries. North America, Australia, and European countries upload the most preprints, with many fewer coming from countries in Africa, Central America, and parts of Asia ([@fig-summary]B).  Articles covered all major taxonomic groups, with the most common groups being vertebrates (`r sum_taxa$perc[6]`%), plants (`r sum_taxa$perc[1]`%), and invertebrates (`r sum_taxa$perc[5]`%) ([@fig-summary]C).

@@

Accepting a greater diversity of article types allows *EcoEvoRxiv* to help deal with the 'grey literature' problem, whereby data that are relevant for research syntheses are not published in typical peer-reviewed journals [@haddaway2015shades; @haddaway2020eight]. *EcoEvoRxiv* has made a concerted effort to diversify the types of articles hosted. This is reflected by `r nt`% of of the articles being books, book chapters, reports, and other research output types, which are typically considered ‘grey literature’ in ecology and evolutionary biology. As a result, articles on *EcoEvoRxiv* are more diverse than those on other preprint servers which have more restrictive submission policies. For example, *bioRxiv* only accepts research articles (https://www.biorxiv.org/submit-a-manuscript).

Empirical research articles are still the most common type of article submitted to *EcoEvoRxiv* (`r sum_types[sum_types$type == "research article",]$perc`%), followed by reviews and meta-analyses (`r sum_types[sum_types$type == "reviews and meta-analyses",]$perc`%) and opinion papers (`r sum_types[sum_types$type == "opinion",]$perc`%) ([@fig-summary]D). Currently, *EcoEvoRxiv* does not host many reports, particularly from government or industry, but has formed fruitful partnerships with the International Union for Conservation of Nature (IUCN). For example, IUCN Red-list Ecosystem Reports are now posted to *EcoEvoRxiv* and our community has been able to work closely with the IUCN to ensure these documents meet the IUCN requirements. We would encourage authors to consider posting books, book chapters, and reports to ensure that they are openly accessible and more easily found. Accepted *EcoEvoRxiv* submissions are given a unique DOI and are indexed on Google Scholar.

@@

Research can take a while to be published (see below). Early and Mid-Career Researchers (EMCRs) (~10 years post-PhD) are under pressure to publish rapidly to be competitive in job applications, promotions, and obtaining grants to progress their careers [@vale2015accelerating; @sarabipour2019value]. Preprints are especially useful for EMCRs because they can achieve faster dissemination and greater visibility [@fu2019releasing]. EMCRs may therefore be expected to make use of preprints more than colleagues at later career stages because they are more often in charge of article submission and have developed their careers in an environment where preprints are a normal part of the publication process. We collected data on the 'academic age' of submitting authors by looking at Google Scholar profiles of authors (when available) and recording their first year of publication in a peer-reviewed journal. While this is a rough estimate of career stage, there was evidence that the number of preprints posted decreases with later career stages (negative binomial glm: year slope = `r res[2, "Estimate"]`, SE: `r round(res[2, "Std. Error"], 2)`, *p* < 0.001, n = `r nrow(count_dat[complete.cases(count_dat),])` years). Most preprints were submitted by authors who published their first paper in the last ~10 years ([@fig-summary]E), with the median year since first publication being `r median(data2$submitting_author_first_publication_year, na.rm = TRUE)` (mean = `r mean(data2$submitting_author_first_publication_year, na.rm = TRUE)`; SD = `r sd(data2$submitting_author_first_publication_year, na.rm = TRUE)`, n = `r sum(complete.cases(data2$submitting_author_first_publication_year))`). These patterns support the expectation that EMCRs may use preprints to make their work more visible and disseminate their findings more quickly. However, we acknowledge that validating this conclusion does require more rigorous experimental approaches.

Gender differences in preprint use and publication outcomes have also been observed in several research fields, including ecology and evolutionary biology [@fox2019gender; @wehner2020comparison]. Therefore, such discrepancies are expected to manifest in preprint use on *EcoEvoRxiv*, but it is unclear to what extent. Understanding gender publishing patterns is challenging with observational data such as ours because we cannot know the gender of authors for certain, but we can use a data-driven approach to ascertain the probability that a particular name is of a given gender (man or woman). To obtain a rough idea of an author's gender, we used the R package *gender* (v.`r utils::packageVersion("gender")`; @mullen2021predict) to predict the most likely gender of the submitting author of a preprint. We only used algorithm-assigned names where the gender was identified with 95% certainty. For the remaining names, we performed manual searches to determine gender based on the pronouns and photographs from professional and personal websites. We acknowledge that our approach does not capture self-assigned and non-binary genders. As such, our assumptions about an author's gender identity may be incorrect. Our data on gender had only two missing values--one where the first name of the submitting author was missing and the other one for a collective submission. As expected, we found that women were less likely to post to *EcoEvoRxiv* compared to men (women: `r (table(data4$gender)[1] / sum(table(data4$gender))*100)`%), which may reflect the broader publishing disparities between male and female scientists [@fox2019gender].

# 3. Following the journey of a preprint on *EcoEvoRxiv*: from submission to publication
```{r, fig-pubsummary}
#| label: fig-pubsummary
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Summary of the publication status of articles on *EcoEvoRxiv*. A) Time between uploading an preprint to *EcoEvoRxiv* and its publication as a peer-reviewed journal article. Preprints were considered those articles that were published a minimum of 2 months (60 days) after being posted to *EcoEvoRxiv*. B) Access status of published articles (both preprints and postprints) linked to *EcoEvoRxiv* preprints classified as “Open Access” or “Not Open Access”. “Unknown” status is for articles not explicitly identified as being open access or not on the Unpaywall platform. C) Sub-types of open access status of published articles on *EcoEvoRxiv*. Sub-type meanings are as follows: ‘Green’, articles published in ‘toll-access journals but achieved in an open access repository; ‘Bronze’, articles are free to read on publishers website without a license but grants no other rights and can be delayed free-to-read; ‘Hybrid’, articles are free to read upon publication with an open access license; ‘Gold’, articles published in fully open access journals. For full details on the meaning of each category see  https://support.unpaywall.org/support/solutions/articles/44001777288-what-do-the-types-of-oa-status-green-gold-hybrid-and-bronze-mean-

@@

We estimated how long it takes to publish a preprint in ecology and evolution by recording the time between when a preprint was first posted on *EcoEvoRxiv*, and its final acceptance in a peer-reviewed journal. In total, `r unpub` submissions remained unpublished (`r (unpub / nrow(data2))*100`%, n = `r nrow(data2)`) at the time when these data were collected. Not all of these preprints, however, are anticipated to be published in a peer-reviewed journal (e.g., reports). Nonetheless, the median time to publication for preprints was `r sum_preprints$median` days (8 months) (mean = `r sum_preprints$mean`; SD = `r sum_preprints$sd` days) with the maximum time to publication being `r sum_preprints$max` days or `r sum_preprints$max/365` years ([@fig-pubsummary]A). Our results largely confirm the extended timeframes that most authors experience between writing their research papers and their publication.
